A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-15 paper.tex
Add sequence-of-trees sentence
Start on eARG section

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
% Recombination (e.g.,
% the cause of the breakpoint at position 7)
% is associated with two parent genomes (\textsf{e} and
% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
% (C) The edge-annotated genome ARG in (B) fully defines the
(C) The corresponding local (``marginal'') trees.
}
\end{figure}

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it useful to have
a concrete definition and process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. The set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

\section*{Event ARGs}\label{eARG}
In keeping with its original derivation in terms of a backward-time stochastic
process (see Appendix~\ref{sec-arg-history}),
the classical definition of an ARG data structure is in terms of evolutionary
\emph{events}.

and we therefore refer to it as an ``event ARG'' (eARG).
In an eARG, nodes represent
common ancestor or recombination events that occurred in the
history of some set of sample genomes, and edges represent ancestral
lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages, and a list of two parents is required.
As illustrated in Figure~\ref{fig-arg-data-structure},
the regions of the genome which are transmitted are usually
specified by associating recombination nodes with a breakpoint.
In other words, eARGs traditionally
carry positional information on their nodes, rather than
storing genome positions as edge annotations.

To recover the local tree at position $x$,
requires traversing the graph rootwards from the leaves.
When we encounter a recombination node during traversal
if $x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent. This means the order in which parents are
listed is important. Alternatively, additional information can be incorporated to distinguish left and right parents \citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.

@@

% Griffiths and colleagues~\citep{griffiths1997ancestral} originally defined
% the ARG as analog of the Kingman coalescent~\citep{kingman1982coalescent},
% in which the ancestry of a sample of $n$ genomes under idealised
% conditions is modelled. One of the key assumptions of this model is that
% the sample size $n$ is much less than the population size $N$, which means
% that the probability of two multiple concurrent events (i.e., coalescence
% and recombination events) is negligible.

@@

\node [anchor=north west] at (20,6) {C};
\node [nodelabel,anchor=north west] at ($(10,5)$) {
\begin{tabular}{ll}
% Edges & Intervals\\
\hline
$(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
$(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
$(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
$(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
$(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
$(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
$(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
\end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A graphical depiction of a classical event ARG (left) and a corresponding encoding of the information (right). For uniformity, rather than defining alternative node types, we
identify recombination nodes (e.g.\ \noderef{d}) as those with a non-null breakpoint; these also have two parents. Nodes \noderef{e}, \noderef{f} and \noderef{g} are
common ancestor events.
}
\end{figure}

@@

% FIXME this needs more work and doesn't really fit into the
% narrative right now, but we do need to make this point.
The key insight that makes the succinct tree sequence encoding
an efficient substrate for defining analysis algorithms is that
it allows us to generate the local trees along the genome
in a way that allows us to reason about the \emph{differences}
between those trees.
Sequentially generating the marginal trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified gARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.

>>>>

\section*{Introduction}
% Inferring the genetic relationships between sampled genomes in the form of a
% phylogenetic tree is a necessary prerequisite for many analyses of
% species~\citep{rannala2003genetics}, non-recombining sections of
% DNA~\citep{cann1987mitochondrial,underhill2001annalsofhumangenetics}
% and viruses~\citep{grenfell2004science}.
% A phylogenetic tree captures information about the genealogy
% of the sample in an elegant and concise
% way, by postulating a set of ancestors common to the samples
% together with paths of inheritance between them. A rich literature exists
% on analysing the mathematical properties of these
% trees~\citep{steel2016phylogeny}, and numerous
% methods exist to infer them~\citep{felsenstein2004inferring}.
% However, the genetic genealogy of a recombining organism
% cannot be captured by a single tree: instead,
% the inheritance of each section of the genome may be described
% by a (slightly) different local tree~\citep{hudson1983properties}.
% Inferring complex
% genealogies of this nature
% presents profound technical difficulties, meaning genealogy-based approaches have
% not been widely used in population and statistical genetics.
% Until recently, analysis of
% population genetic data has therefore primarily focussed on summaries derived
% from the local trees~\citep{tajima1983evolutionary,tavare1984line}
% rather than using the full underlying genealogy itself. Such
% derived measures include allele frequencies and site frequency
% spectra~\citep{achaz2009frequency,ralph2020efficiently},
% patterns of linkage disequilibrium~\citep{mcvean2002genealogical}, and
% principal components~\citep{mcvean2009genealogical}.
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high-level, ARGs describe the sequence of correlated genealogical
trees along the genome, induced by the different paths of
genetic inheritance caused by recombination.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An example gARG embedded in a pedigree.
(A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal and maternal  genomes
labelled \textsf{a} to \textsf{p}. Black lines show inheritance paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Genomes \textsf{a} and \textsf{c} are the product of two independent
recombination events between
the paternal genomes \textsf{e}
and \textsf{f}, and regions of genome inherited are shown.
Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker.
(B) The corresponding gARG along with inheritance annotations on all edges
(partial inheritance in bold).
% Recombination (e.g.,
% the cause of the breakpoint at position 7)
% is associated with two parent genomes (\textsf{e} and
% \textsf{f}) and a child genome (the recombinant, \textsf{c}).
% (C) The edge-annotated genome ARG in (B) fully defines the
(C) The corresponding local trees.
}
\end{figure}

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local
% TODO do we mention "marginal" anywhere? Might not be worth the
% effort, surely "local" is clear?
is somewhat subtle (see section XXX), and it useful to have
a concrete definition and process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. The set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs model the inheritance between specific genomes,
annotating the edges of the graph with the intervals
of genetic inheritance. It is primarily concerned with
describing relationships and the details of inheritance
between pairs of genomes.
In contrast, the classical ARG data structure is defined
by evolutionary \emph{events},
and we therefore refer to it as an ``event ARG'' (eARG).
In an eARG we usually have two different types
of node,
representing common ancestor and recombination events that occurred in the
history of some set of sample genomes
and edges represent ancestral lineages~\citep{griffiths1996ancestral}.
In a common ancestor event the inbound lineages are merged into a
single ancestral lineage with one parent, whereas in a recombination
event a lineage is split into two independent
ancestral lineages with an associated breakpoint.
Figure~\ref{fig-arg-data-structure} shows an example of a classical
eARG with three samples (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event, along with an example concrete
encoding of the data.  There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
We use the standard convention here
that the order of parents
is significant, but additional information can be incorporated to
distinguish left and right parents
\citep[e.g.][Chap.\ 3]{gusfield2014recombinatorics}.

Beyone the interpretation of nodes as representing events or
genomes (which may reasonably be seen as somewhat philosophical),
the essential difference between a gARG and a classical eARG
[JK WIP from here:]
is that eARGs define the

carry positional information on their nodes, rather than
storing genome positions as edge annotations.

the regions of the genome which are transmitted are usually
specified by associating recombination nodes with a breakpoint.
In other words, eARGs
carry positional information on their nodes, rather than
storing genome positions as edge annotations.

To recover the local tree at position $x$,
requires traversing the graph rootwards from the leaves.
When we encounter a recombination node during traversal
if $x$ is less than the breakpoint we follow
the edge to the first parent, and otherwise follow the edge
to the second parent. This means the order in which parents are
listed is important.

@@

@@

% \node [anchor=north west] at (20,6) {C};
% \node [nodelabel,anchor=north west] at ($(10,5)$) {
% \begin{tabular}{ll}
% % Edges & Intervals\\
% \hline
% $(\textsf{a}, \textsf{e})$ & $[0, L)$ \\
% $(\textsf{b}, \textsf{d})$ & $[0, L)$ \\
% $(\textsf{c}, \textsf{f})$ & $[0, L)$ \\
% $(\textsf{d}, \textsf{e})$ & $[0, x)$ \\
% $(\textsf{d}, \textsf{f})$ & $[x, L)$ \\
% $(\textsf{e}, \textsf{g})$ & $[0, L)$ \\
% $(\textsf{f}, \textsf{g})$ & $[0, L)$ \\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
A graphical depiction of a classical event ARG (left) and a corresponding
encoding of the information (right).
Recombination nodes (e.g.\ \noderef{d}) have a non-null breakpoint
and two parents.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
}
\end{figure}

@@

% FIXME this needs more work and doesn't really fit into the
% narrative right now, but we do need to make this point.
The key insight that makes the succinct tree sequence encoding
an efficient substrate for defining analysis algorithms is that
it allows us to generate the local trees along the genome
in a way that allows us to reason about the \emph{differences}
between those trees.
Sequentially generating the local trees along the genome
is also fundamental, and is necessary whenever we need to
perform calculations that are contingent on more than just the
isolated properties of an edge. \cite{kelleher2016efficient}
showed how all trees can be sequentially generated in
constant time per tree transition in a fully simplified gARG.
Furthermore, we can easily reason about how tree topologies
change (and stay the same), leading to efficient algorithms
for computing population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently},
implementing the Li and Stephens
model~\citep{kelleher2019inferring,wohns2022unified}
and likely many more.
